---
title: "Spatial visualization for multiplexed imaging cytometry data in R"
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('cytomapper')`"
author:
- name: Nils Eling
  affiliation: Department for Quantitative Biomedicine, University of Zurich
  email: nils.eling@dqbm.uzh.ch.com
- name: Nicolas Damond
  affiliation: Department for Quantitative Biomedicine, University of Zurich
  email: nicolas.damond@dqbm.uzh.ch.com
output:
  BiocStyle::html_document:
    toc_float: yes
bibliography: library.bib
abstract: |
    Highly multiplexed imaging cytometry acquires the single-cell expression of
    selected proteins in a spatially-resolved fashion. These measurements can be
    visualized across multiple length-scales. First, pixel-level intensities
    represent the spatial distributions of feature expression with highest
    resolution. Second, after segmentation, expression values or cell-level
    metadata (e.g. cell-type information) can be visualized on segmented cell
    areas. This package contains functions for the visualization of multiplexed
    read-outs and cell-level information obtained by multiplexed imaging
    cytometry. The main functions of this package allow 1. the visualization of
    pixel-level information across multiple channels and 2. the display of
    cell-level information (expression and/or metadata) on segmentation masks.
vignette: |
  %\VignetteIndexEntry{Spatial visualization for multiplexed imaging cytometry data in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE)
library(BiocStyle)
```

```{r library, echo=FALSE}
library(cytomapper)
```

# Introduction

This vignette gives an introduction to displaying highly multiplexed imaging cytometry data with the `cytomapper` package.
As an example, these instructions display imaging mass cytometry (IMC) data.
However, other imaging cytometry approaches including multiplexed ion beam imaging (MIBI) [@angelo2014mibi], tissue-based cyclic immunofluorescence (t-CyCIF) [@lin2018cycif] and iterative indirect immunofluorescence imaging (4i) [@gut20184i], which produce pixel-level intensities and optionally segmentation masks can be displayed using `cytomapper`.

IMC [@giesen2014imc] is a multiplexed imaging approach to measure spatial protein abundance.
In IMC, tissue sections are stained with a mix of $\tilde{}40$ metal-conjugated antibodies prior to laser ablation with $1\mu{}m$ resolution. 
The ablated material is transferred to a mass cytometer for time-of-flight detection of the metal ions [@giesen2014imc][@mavropoulosimc]. 
In that way, hundreds of images (usually with an image size of around 1mm x 1mm) can be generated in a reasonable amount of time [@damond2019pancreas]. 

Raw IMC data are computationally processed using a segmentation pipeline (available at [https://github.com/BodenmillerGroup/ImcSegmentationPipeline](https://github.com/BodenmillerGroup/ImcSegmentationPipeline)). 
This pipeline produces image stacks containing the raw pixel values for up to 40 channels, segmentation masks containing the segmented cells, cell-level expression and metadata information as well as a number of image-level meta information.

Cell-level expression and metadata can be processed and read into a `SingleCellExperiment` class object.
For more information on the `SingleCellExperiment` object and how to create it, please see the `r Biocpkg("SingleCellExperiment")` package and the [Orchestrating Single-Cell Analysis with Bioconductor](https://osca.bioconductor.org/data-infrastructure.html#the-singlecellexperiment-class) workflow.

The `cytomapper` package provides a new `CytoImageList` class as a container for multiplexed images or segmentation masks.
For more information on this class, refer to the [CytoImageList](#CytoImageList) section.

The main functions of this package include `plotCells` and `plotPixels`.
The `plotCells` function requires the following object inputs to display cell-level information (expression and metadata):

1. a SingleCellExperiment object, which contains the cells' counts and metadata
2. an CytoImageList object containing the segmentation mask per image

The `plotPixels` function requires the following object inputs to display pixel-level expression information:

1. a `CytoImageList` object containing the pixel-level information per channel
2. (optionally) a `SingleCellExperiment` object, which contains the cells' counts and metadata
2. (optionally) a `CytoImageList` object containing the segmentation mask per image

# Quick start

The following section provides a quick example highlighting the functionality of `cytomapper`.
For detailed information on reading in the data, refer to the [Reading in data](#ReadingData) section.
More information on the required data format is provided in the [Data formats](#DataFormats) section.
In the first step, we will read in the provided [toy dataset](#ToyData)

```{r quickstart-load-data}
data(pancreasSCE)
data(pancreasImages)
data(pancreasMasks)
```

The `CytomImageList` object containing pixel-level intensities representing the ion counts for five proteins can be displayed using the `plotPixel` function:

```{r quickstart-plotPixels}
plotPixels(image = pancreasImages, colour_by = c("H3", "SMA", "CD44"))
```

For more details on image normalization, cell outlining, and other pixel-level manipulations, refer to the [Plotting pixel information](#PixelInfo) section.

The `CytoImageList` object containing segmentation masks that represent the cell areas on the image can be displayed using the `plotCells` function.
Only the segmentation masks are plotted when no other parameters are specified.

```{r quickstart-plotCells}
plotCells(mask = pancreasMasks)
```

To colour and/or outline segmentation masks, a `SingleCellExperiment`, an `img_id` and `cell_id` entry need to be specified:

```{r quickstart-plotCells-2}
plotCells(mask = pancreasMasks, object = pancreasSCE, cell_id = "CellNb", img_id = "ImageNb",
          colour_by = "SMA")
plotCells(mask = pancreasMasks, object = pancreasSCE, cell_id = "CellNb", img_id = "ImageNb",
          colour_by = "SMA", outline_by = "CellType")
plotCells(mask = pancreasMasks, object = pancreasSCE, cell_id = "CellNb", img_id = "ImageNb", 
          colour_by = "CellType")
```

For more information on the data formats and requirements, refer to the following section.
More details on the `plotCells` function are provided in the [Plotting cell information](#CellInfo) section.

# Data formats {#DataFormats}

The `cytomapper` package combines objects of the `r Biocpkg("SingleCellExperiment")` class and the `CytoImageList` class (provided in `cytomapper`) to visualize cell- and pixel-level information.

In the main functions of the package, `image` refers to an `CytoImageList` object containing one or multiple multi-channel images where each channel represents the pixel-intensity of one selected marker (proteins in the case of IMC).
The entry `mask` refers to an `CytoImageList` object containing one or multiple segmentation masks.
Segmentation masks are defined as one-channel images containing integer values, which represent the cells' ids or 0 (background).
Finally, the `object` entry refers to a `SingleCellExperiment` class object that contains cell-specific expression values (in the `assay` slots) and cell-specific metadata in the `colData` slot.

To link information between the `SingleCellExperiment` and `CytoImageList` objects, two slots need to be specified:

* `img_id`: a single character indicating the `colData` (in the `SingleCellExperiment` object) and `elementMetadata` (in the `CytoImageList` object) entry that contains the image identifiers. These image ids have to match between the `SingleCellExperiment` object and the `CytoImageList` object.
* `cell_id`: a single character indicating the `colData` entry that contains the cell identifiers. These should be integer values corresponing to pixel-values in the segmentation masks.

The `img_id` and `cell_id` entry in the `SingleCellExperiment` object need to be accessible via: 

```{r image-cell-id-sce}
colData(pancreasSCE)[,"ImageNb"]
colData(pancreasSCE)[,"CellNb"]
```

The `img_id` entry in the `CytoImageList` object need to be accessible via: 

```{r image-cell-id-cil}
mcols(pancreasImages)[,"ImageNb"]
mcols(pancreasMasks)[,"ImageNb"]
```

For more information on the `CytoImageList` class, please refer to the section [The CytoImageList object](#CytoImageList).
For more information on the `SingleCellExperiment` object and how to create it, please see the `r Biocpkg("SingleCellExperiment")` package and the [Orchestrating Single-Cell Analysis with Bioconductor](https://osca.bioconductor.org/data-infrastructure.html#the-singlecellexperiment-class) workflow.

## The provided toy dataset {#ToyData}

For visualization purposes, the `cytomapper` package provides a toy dataset containing 3 images of $100\mu{m}$ x $100\mu{m}$ dimensions (100 x 100 pixels).
The dataset contains 282 segmented cells and the expression values for 5 proteins: H3, SMA, INS, CD38, and CD44
It represents a small subset of the data presented in [A Map of Human Type 1 Diabetes Progression by Imaging Mass Cytometry](Ahttps://www.cell.com/cell-metabolism/fulltext/S1550-4131(18)30691-0?_returnURL=https%3A%2F%2Flinkinghub.elsevier.com%2Fretrieve%2Fpii%2FS1550413118306910%3Fshowall%3Dtrue).

Pixel-level intensities for all 5 markers (5 channels) are stored in the `pancreasImages` object.
Entires to the `CytoImageList` object and the rownames of the object match: A02_imc, D01_imc, and F01_imc
The `elementMetadata` slot (accesible via the `mcols()` function) contains the image identifiers.

```{r pancreasImages}
pancreasImages
mcols(pancreasImages)
channelNames(pancreasImages)
imageData(pancreasImages[[1]])[1:15,1:5,1]
```

The corresponding segmentation masks are stored in the `pancreasMasks` object.
Segmentation masks are defined as one-channel images containing integer values, which represent the cells' ids or 0 (background).

```{r pancreasMasks}
pancreasMasks
mcols(pancreasMasks)
imageData(pancreasMasks[[1]])[1:15,1:5]
```

All cell-specific metadata are stored in the `colData` slot of the corresponding `SingleCellExperiment` object: `pancreasSCE`.
For more information on the metadata, please refer to the `?pancreasSCE` documentation.
Of note: the cell-type labels contained in the `colData(pancreasSCE)$CellType` slot are randomized and do not represent biologically relevant cell-types.

```{r pancreasSCE}
pancreasSCE
names(colData(pancreasSCE))
```

The `pancreasSCE` object also contains further information on the measured proteins via the `rowData(pancreasSCE)` slot.
Furthermore, the `pancreasSCE` object contains the raw expression counts per cell in the form of mean pixel value per cell and protein (accessible via `counts(pancreasSCE)`).
The arcsinh-transformed (using a co-factor of 1) raw expression counts can be obtained via `assay(pancreasSCE, "exprs")`.

# Reading in data {#ReadingData}

The `cytomapper` package provides the `loadImages` function to conveniently read images into a `CytoImageList` object.

## Load images



Introduction to the `LoadImage` function

## Add metadata
```{r}
mcols(pancreasImages)
mcols(pancreasMasks)
colData(pancreasSCE)$ImageNb
```


# The CytoImageList object {#CytoImageList}

This document also introduces the `CytoImageList` class, which inherits from the [`SimpleList` class](https://rdrr.io/bioc/S4Vectors/man/SimpleList-class.html).

Images should be stacks, all stacks should have the same number of channels
Masks should be single-channel images

Explain mcols

## Accessors

## Coercion

## Subsetting
Extract one channel from the CytoImageList

## Looping

# Plotting pixel information {#PixelInfo}

## Outline

## Normalization

# Plotting cell information {#CellInfo}

# Customization

## Legend

## Scale bar

## Cropping

# Saving images

# Acknowledgements

# Contributions

Nicolas created cytomapper - Nicolas and Nils implemented the package ...

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

# References

