---
title: "Visualization of imaging cytometry data in R"
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('cytomapper')`"
author:
- name: Nils Eling
  affiliation: Department for Quantitative Biomedicine, University of Zurich
  email: nils.eling@dqbm.uzh.ch.com
- name: Nicolas Damond
  affiliation: Department for Quantitative Biomedicine, University of Zurich
  email: nicolas.damond@dqbm.uzh.ch.com
output:
    BiocStyle::html_document:
        toc_float: yes
bibliography: library.bib
abstract: |
    Highly multiplexed imaging cytometry acquires the single-cell expression of
    selected proteins in a spatially-resolved fashion. These measurements can be
    visualized across multiple length-scales. First, pixel-level intensities
    represent the spatial distributions of feature expression with highest
    resolution. Second, after segmentation, expression values or cell-level
    metadata (e.g. cell-type information) can be visualized on segmented cell
    areas. This package contains functions for the visualization of multiplexed
    read-outs and cell-level information obtained by multiplexed imaging
    cytometry. The main functions of this package allow 1. the visualization of
    pixel-level information across multiple channels and 2. the display of
    cell-level information (expression and/or metadata) on segmentation masks.
vignette: |
    %\VignetteIndexEntry{"Visualization of imaging cytometry data in R}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE,
                        fig.retina = 0.75, crop = NULL)
library(BiocStyle)
```

```{r library, echo=FALSE}
library(cytomapper)
```

# Introduction

This vignette gives an introduction to displaying highly multiplexed imaging
cytometry data with the `cytomapper` package. As an example, these instructions
display imaging mass cytometry (IMC) data. However, other imaging cytometry
approaches including multiplexed ion beam imaging (MIBI) [@angelo2014mibi],
tissue-based cyclic immunofluorescence (t-CyCIF) [@lin2018cycif] and iterative
indirect immunofluorescence imaging (4i) [@gut20184i], which produce pixel-level
intensities and optionally segmentation masks can be displayed using
`cytomapper`.

IMC [@giesen2014imc] is a multiplexed imaging approach to measure spatial
protein abundance. In IMC, tissue sections are stained with a mix of
$\tilde{}40$ metal-conjugated antibodies prior to laser ablation with $1\mu{}m$
resolution. The ablated material is transferred to a mass cytometer for
time-of-flight detection of the metal ions [@giesen2014imc][@mavropoulosimc]. In
that way, hundreds of images (usually with an image size of around 1mm x 1mm)
can be generated in a reasonable amount of time [@damond2019pancreas].

Raw IMC data are computationally processed using a segmentation pipeline
(available at
[https://github.com/BodenmillerGroup/ImcSegmentationPipeline](https://github.com/BodenmillerGroup/ImcSegmentationPipeline)).
This pipeline produces image stacks containing the raw pixel values for up to 40
channels, segmentation masks containing the segmented cells, cell-level
expression and metadata information as well as a number of image-level meta
information.

Cell-level expression and metadata can be processed and read into a
`SingleCellExperiment` class object. For more information on the
`SingleCellExperiment` object and how to create it, please see the `r
Biocpkg("SingleCellExperiment")` package and the [Orchestrating Single-Cell
Analysis with
Bioconductor](https://osca.bioconductor.org/data-infrastructure.html#the-singlecellexperiment-class)
workflow.

The `cytomapper` package provides a new `CytoImageList` class as a container for
multiplexed images or segmentation masks. For more information on this class,
refer to the [CytoImageList](#CytoImageList) section.

The main functions of this package include `plotCells` and `plotPixels`. The
`plotCells` function requires the following object inputs to display cell-level
information (expression and metadata):

1. a SingleCellExperiment object, which contains the cells' counts and metadata
2. a CytoImageList object containing the segmentation mask per image

The `plotPixels` function requires the following object inputs to display
pixel-level expression information:

1. a `CytoImageList` object containing the pixel-level information per channel
2. (optionally) a `SingleCellExperiment` object, which contains the cells'
counts and metadata 2. (optionally) a `CytoImageList` object containing the
segmentation mask per image

# Quick start {#QuickStart}

The following section provides a quick example highlighting the functionality of
`cytomapper`. For detailed information on reading in the data, refer to the
[Reading in data](#ReadingData) section. More information on the required data
format is provided in the [Data formats](#DataFormats) section. In the first
step, we will read in the provided [toy dataset](#ToyData)

```{r quickstart-load-data}
data(pancreasSCE)
data(pancreasImages)
data(pancreasMasks)
```

The `CytoImageList` object containing pixel-level intensities representing the
ion counts for five proteins can be displayed using the `plotPixel` function:

```{r quickstart-plotPixels}
plotPixels(image = pancreasImages, colour_by = c("H3", "SMA", "CD44"))
```

For more details on image normalization, cell outlining, and other pixel-level
manipulations, refer to the [Plotting pixel information](#PixelInfo) section.

The `CytoImageList` object containing segmentation masks that represent the cell
areas on the image can be displayed using the `plotCells` function. Only the
segmentation masks are plotted when no other parameters are specified.

To colour and/or outline segmentation masks, a `SingleCellExperiment`, an
`img_id` and `cell_id` entry need to be specified:

```{r quickstart-plotCells-2}
plotCells(mask = pancreasMasks, object = pancreasSCE, 
            cell_id = "CellNb", img_id = "ImageNb", colour_by = "SMA",
            outline_by = "CellType")
plotCells(mask = pancreasMasks, object = pancreasSCE, 
            cell_id = "CellNb", img_id = "ImageNb", colour_by = "CellType")
```

For more information on the data formats and requirements, refer to the
following section. More details on the `plotCells` function are provided in the
[Plotting cell information](#CellInfo) section.

# Data formats {#DataFormats}

The `cytomapper` package combines objects of the `r
Biocpkg("SingleCellExperiment")` class and the `CytoImageList` class (provided
in `cytomapper`) to visualize cell- and pixel-level information.

In the main functions of the package, `image` refers to a `CytoImageList` object
containing one or multiple multi-channel images where each channel represents
the pixel-intensity of one selected marker (proteins in the case of IMC). The
entry `mask` refers to a `CytoImageList` object containing one or multiple
segmentation masks. Segmentation masks are defined as one-channel images
containing integer values, which represent the cells' ids or 0 (background).
Finally, the `object` entry refers to a `SingleCellExperiment` class object that
contains cell-specific expression values (in the `assay` slots) and
cell-specific metadata in the `colData` slot.

To link information between the `SingleCellExperiment` and `CytoImageList`
objects, two slots need to be specified:

* `img_id`: a single character indicating the `colData` (in the
`SingleCellExperiment` object) and `elementMetadata` (in the `CytoImageList`
object) entry that contains the image identifiers. These image ids have to match
between the `SingleCellExperiment` object and the `CytoImageList` object. 
* `cell_id`: a single character indicating the `colData` entry that contains the
cell identifiers. **These should be integer values corresponding to pixel-values
in the segmentation masks**.

The `img_id` and `cell_id` entry in the `SingleCellExperiment` object need to be
accessible via:

```{r image-cell-id-sce}
head(colData(pancreasSCE)[,"ImageNb"])
head(colData(pancreasSCE)[,"CellNb"])
```

The `img_id` entry in the `CytoImageList` object need to be accessible via: 

```{r image-cell-id-cil}
mcols(pancreasImages)[,"ImageNb"]
mcols(pancreasMasks)[,"ImageNb"]
```

For more information on the `CytoImageList` class, please refer to the section
[The CytoImageList object](#CytoImageList). For more information on the
`SingleCellExperiment` object and how to create it, please see the `r
Biocpkg("SingleCellExperiment")` package and the [Orchestrating Single-Cell
Analysis with
Bioconductor](https://osca.bioconductor.org/data-infrastructure.html#the-singlecellexperiment-class)
workflow.

## The provided toy dataset {#ToyData}

For visualization purposes, the `cytomapper` package provides a toy dataset
containing 3 images of $100\mu{m}$ x $100\mu{m}$ dimensions (100 x 100 pixels).
The dataset contains 282 segmented cells and the expression values for 5
proteins: H3, SMA, INS, CD38, and CD44 It represents a small subset of the data
presented in [A Map of Human Type 1 Diabetes Progression by Imaging Mass
Cytometry](Ahttps://www.cell.com/cell-metabolism/fulltext/S1550-4131(18)30691-0?_returnURL=https%3A%2F%2Flinkinghub.elsevier.com%2Fretrieve%2Fpii%2FS1550413118306910%3Fshowall%3Dtrue).

Pixel-level intensities for all 5 markers (5 channels) are stored in the
`pancreasImages` object. Entires to the `CytoImageList` object and the rownames
of the object match: A02_imc, D01_imc, and F01_imc The `elementMetadata` slot
(accesible via the `mcols()` function) contains the image identifiers.

```{r pancreasImages}
pancreasImages
mcols(pancreasImages)
channelNames(pancreasImages)
imageData(pancreasImages[[1]])[1:15,1:5,1]
```

The corresponding segmentation masks are stored in the `pancreasMasks` object.
Segmentation masks are defined as one-channel images containing integer values,
which represent the cells' ids or 0 (background).

```{r pancreasMasks}
pancreasMasks
mcols(pancreasMasks)
imageData(pancreasMasks[[1]])[1:15,1:5]
```

All cell-specific metadata are stored in the `colData` slot of the corresponding
`SingleCellExperiment` object: `pancreasSCE`. For more information on the
metadata, please refer to the `?pancreasSCE` documentation. Of note: the
cell-type labels contained in the `colData(pancreasSCE)$CellType` slot are
randomized and do not represent biologically relevant cell-types.

```{r pancreasSCE}
pancreasSCE
names(colData(pancreasSCE))
```

The `pancreasSCE` object also contains further information on the measured
proteins via the `rowData(pancreasSCE)` slot. Furthermore, the `pancreasSCE`
object contains the raw expression counts per cell in the form of mean pixel
value per cell and protein (accessible via `counts(pancreasSCE)`). The
arcsinh-transformed (using a co-factor of 1) raw expression counts can be
obtained via `assay(pancreasSCE, "exprs")`.

# Reading in data {#ReadingData}

The `cytomapper` package provides the `loadImages` function to conveniently read
images into a `CytoImageList` object.

## Load images

The `loadImages` function returns a `CytoImageList` object containing the
multi-channel images or segmentation masks. Refer to the `?loadImages` function
to see the full functionality.

As an example, we will read in multi-channel images and segmentation masks
provided by the `cytomapper` package.

```{r read-in-images}
# Read in masks
path.to.images <- system.file("extdata", package = "cytomapper")
all_masks <- loadImages(path.to.images, pattern = "_mask.tiff")
all_masks

# Read in images
all_stacks <- loadImages(path.to.images, pattern = "_imc.tiff")
all_stacks
```

## Add metadata

To link images between the two `CytoImageList` objects and the corresponding
`SingleCellExperiment` object, the image ids need to be added to the
`elementMetadata` slot of the `CytoImageList` objects. From the experimental
setup, we know that the image named 'A02_imc' has image id '1', D01_imc has id
'2', F01_imc has id '3'.

```{r add-metadata}
unique(pancreasSCE$ImageNb)
mcols(all_masks)$ImageNb <- c("1", "2", "3")
mcols(all_stacks)$ImageNb <- c("1", "2", "3")
```

## Scale images

We can see that, in some cases, the pixel-values are not correctly scaled by the
image encoding. The segmentation masks should only contain integer entries:

```{r image-encoding}
head(unique(as.numeric(all_masks[[1]])))
```

The provided data was processed using CellProfiler [@carpenter2006cellprofiler].
By default, CellProfiler scales all pixel intensities between 0 and 1. This is
done by dividing each count by the maximum possible intensity value (see
[MeasureObjectIntensity](https://cellprofiler-manual.s3.amazonaws.com/CellProfiler-3.0.0/modules/measurement.html#measureobjectintensity)
for more info). In the case of 16-bit encoding (where 0 is a valid intensity),
this scaling value is `2^16-1 = 65535`. Therefore, pixel-intensites need to be
rescaled by this value. However, this scaling value can change and different
images can be scaled by different factors. The user needs make sure to select
the correct factors in more complex cases.

The `cytomapper` package provides a `?scaleImages` function. The user needs to
manually scale images to obtain the correct pixel-values. Here, we scale the
segmentation masks by the factor for 16-bit encoding: `2^16-1`

```{r scaleImage-1}
all_masks <- scaleImages(all_masks, 2^16-1)
head(unique(as.numeric(all_masks[[1]])))
```

In this case, the multi-channel images are not affected by this scaling factor.

## Add channel names

To access the correct images in the multi-channel `CytoImageList` object, the
user needs to set the correct channel names. For this, the `cytomapper` package
provides the `?channelNames` getter and setter function:

```{r channelNames-example}
channelNames(all_stacks) <- c("H3", "SMA", "INS", "CD38", "CD44")
```

The read-in data can now be used for visualization as explained in the [Quick
start section](#QuickStart).

# The CytoImageList object {#CytoImageList}

The `cytomapper` package provides a new `CytoImageList` class, which inherits
from the [`SimpleList`
class](https://rdrr.io/bioc/S4Vectors/man/SimpleList-class.html). Each entry to
the `CytoImageList` object is an `Image` class object defined in the `r
BiocStyle::Biocpkg("EBImage")` package. A `CytoImageList` object is restricted
to the following entries:

* all images need to have the same number of channels
* the order/naming of channels need to be the same across all images
* entries to the `CytoImageList` object need to be uniquely named
* names of `CytoImageList` object can either be `NULL` or should not contain
`NA` or empty entries
* only grayscale images are supported (see `?Image` for more information)
* channels names do not support duplicated entries

`CytoImageList` objects that contain masks should only contain a single channel

The following paragraphs will explain further details on manipulating
`CytoImageList` objects

## Accessors

All accessor functions defined for `SimpleList` also work on `CytoImageList`
class objects. Element-wise metadata --- in the case of the `CytoImageList`
object these are image-specific metadata --- are saved in the `elementMetadata`
slot. This slot can be accessed via the `mcols()` function:

```{r mcols}
mcols(pancreasImages)
mcols(pancreasImages)$PatientID <- c("Patient1", "Patient2", "Patient3")
mcols(pancreasImages)
```

Subsetting a `CytoImageList` object works similar to a `SimpleList` object:

```{r subsetting-SimpleList}
pancreasImages[1]
pancreasImages[[1]]
```

However, to facilitate subsetting and making sure that entry names are
transfered between objects, the `cytomapper` package provides a number of getter
and setter functions:

### Getting and setting images

Individual or multiple entries in a `CytoImageList` object can be obtained or
replaced using the `getImages` and `setImages` functions, respectively.

```{r get-set-images}
cur_image <- getImages(pancreasImages, "A02_imc")
cur_image
setImages(pancreasImages, "New_image") <- cur_image
pancreasImages
mcols(pancreasImages)
```

The `setImages` function ensures that names are transfered from one to the other
object along the assignment operator:

```{r set-images}
names(cur_image) <- "Replacement"
setImages(pancreasImages, 2) <- cur_image
pancreasImages
mcols(pancreasImages)
```

However, if the image to replace is called by name, only the image and
associated metadata is replaced:

```{r set-images-2}
setImages(pancreasImages, "F01_imc") <- cur_image
pancreasImages
mcols(pancreasImages)
```

Images can be deleted by setting the entry to `NULL`:

```{r deleting-images}
setImages(pancreasImages, c("Replacement", "New_image")) <- NULL
pancreasImages
```

**Of note:** for plotting, the entries in the `img_id` slot in the
`CytoImageList` objects have to be unique.

### Getting and setting channels

The `cytomapper` package also provides functions to obtain and replace channels.
This functionality is provided via the `getChannels` and `setChannels`
functions:

```{r get-set-channels}
cur_channel <- getChannels(pancreasImages, "H3")
cur_channel
channelNames(cur_channel) <- "New_H3"
setChannels(pancreasImages, 1) <- cur_channel
pancreasImages
```

The `setChannels` function does not allow combining and adding new channels. For
this task, the `cytomapper` package provides the `mergeChannels` section in the
next paragraph.

### Naming and merging channels

Channel names can be obtained and replaced using the `channelNames` getter and
setter function:

```{r channelnames}
channelNames(pancreasImages)
channelNames(pancreasImages) <- c("ch1", "ch2", "ch3", "ch4", "ch5")
pancreasImages
```

Furthermore, channels can be merged using the `mergeChannels` function:

```{r mergechannels}
cur_channels <- getChannels(pancreasImages, 1:2)
channelNames(cur_channels) <- c("new_ch1", "new_ch2")
pancreasImages <- mergeChannels(pancreasImages, cur_channels) 
pancreasImages
```

## Looping

To perform custom operations on each individual entry to a `CytoImageList`
object, the `r Biocpkg("S4Vectors")` package provides the `endoapply` function.
While the `lapply` function returns a `list` object, the `endoapply` function
provides an object of the same class of the input object.

This allows the user to apply all functions provided by the `r
Biocpkg("EBImage")` package to individual entries within the `CytoImageList`
object:

```{r endoapply-example}
data("pancreasImages")

# Performing a gaussian blur
pancreasImages <- endoapply(pancreasImages, gblur, sigma = 1)
```

# Plotting pixel information {#PixelInfo}

The `cytomapper` package provides the `plotPixels` function to plot pixel-level
intensities of marker proteins. The function requires a `CytoImageList` object
containing a single or multiple multi-channel images. To colour images based on
channel name, the `channelNames` of the object need to be set. Furthermore, to
outline cells, a `CytoImageList` object containing segmentation masks and a
`SingleCellExperiment` object containing cell-specific metadata need to be
provided.

By default, pixel values are coloured internally and scaled between the minimum
and maximum values across all displayed images. However, to manipulate pixel
values and to linearly scale values to a certain range, the `cytomapper` package
provides a function for image normalization.

## Normalization

The `normalize` function provided in the `cytomapper` package internally calls
the `normalize` function of the `r Biocpkg("EBImage")` package. The main
difference between the two functions is the option to scale per image or
globally in the `cytomapper` package. Furthermore, the `cytomapper` package
`normalize` function (see `?'normalize,CytoImageList-method'`) allows the user
to define a clipping range based on percentiles.

By default, the `normalize` function linearly scales the images channel-wise
across all images and returns values between 0 and 1:

```{r normalize-default}
data("pancreasImages")

# Default normalization
cur_images <- normalize(pancreasImages)
```

A `CytoImageList` object can also be normalized image-wise:

```{r normalize-image-wise}
# Image-wise normalization
cur_images <- normalize(pancreasImages, separateImages = TRUE)
```

To clip the image range, the user can provide a percentage-based clipping range:

```{r normalize-clippingrange}
# Percentage-based clipping range
cur_images <- normalize(pancreasImages, percentileRange = c(0,0.99))
plotPixels(cur_images, colour_by = c("H3", "SMA", "CD44"))
```

For more information on the normalization functionality provided by the
`cytomapper` package, please refer to `?'normalize,CytoImageList-method'`.

## Colouring

The `cytomapper` package supports the visualization of up to 6 channels and
displays a combined image by setting the `colour_by` parameter.
See `?plotPixels` for examples.

## Adjusting brightness, contrast and gamma

To enhance individual channels, the brightness (b), contrast (c) and gamma (g)
can be set channel-wise via the `bcg` parameter. These parameters are set in
form of a named `list` object. Entry names need to correspond by channels
specified in `colour_by`. Each entry takes a numeric vector of length three
where the first entry represents the brightness value, the second the contrast
factor and the third the gamma factor. Internally, the brightness value is added
to each channel; each channel is multiplied by the contrast factor and each
channel is exponentiated by the gamma factor.

```{r bcg-pixels}
data("pancreasImages")
# Increase contrast for the SMA and CD44 channel
plotPixels(pancreasImages, 
            colour_by = c("H3", "SMA", "CD44"), 
            bcg = list(SMA = c(0,2,1), 
                        CD44 = c(0,2,1)))
```

## Outlining {#Outlining}

The cells can be outlined when providing a `CytoImageList` object containing the
corresponding segmentation masks and a character `img_id` indicating the name of
the `elementMetadata` slot that contains the image IDs.

The user can furthermore specify the metadata entry to outline cells by. For
this, a `SingleCellExperiment` object containing the cell-specific metadata and
a `cell_id` indicating the name of the `colData` slot that contains the cell IDs
need to be provided:

```{r outline-1-pixels}
plotPixels(pancreasImages, mask = pancreasMasks, 
            object = pancreasSCE, img_id = "ImageNb", 
            cell_id = "CellNb",
            colour_by = c("H3", "SMA", "CD44"),
            outline_by = "CellType")
```

## Subsetting {#subsetting}

The user can further subset the images internally for visualization by setting
the `subset_images` parameter. Images can be subsetted by image ID, entry name
or simply by indexing. In the toy dataset, image IDs are numeric entries and
therefore the `subset_images` parameter behaves the same for image ID subsetting
and subsetting via integer indexing.

```{r subsetting-pixels}
plotPixels(pancreasImages, colour_by = c("H3", "SMA", "CD44"),
            subset_images = "F01_imc")
```

## Adjusting the colour

The user can also customize the colours for selected features. The `colour`
parameter takes a named `list` in which names correspond to the entries to
`colour_by`. To colour continous features such as expression or continous
metadata entries (e.g. cell area, see next section), at least two colours for
interpolation need to be provided. These colours are passed to the
`colorRampPalette` function for interpolation.
For details, please refer to the next [Adjusting the colour](#adjustingColour)
section

# Plotting cell information {#CellInfo}

In the following sections, the `plotCells` function will be introduced. This
function displays cell-level information on segmentation masks. It requires a
`CytoImageList` object containing segmentation masks in the form of
single-channel images. Furthermore, to colour and outline cells, a
`SingleCellExperiment` object containing cell-specific expression counts and
metadata needs to be provided.

By default, cell-specific expression values are coloured internally and scaled
marker-specifically between the minimum and maximum values across the full
`SingleCellExperiment`.

## Colouring

Segmentation masks can be coloured based on the pixel-values averaged across the
area of each cell. In the `SingleCellExperiment` object, these values can be
obtained from the `counts()` slot. To colour segmentation masks based on
expression, the `rownames` of the `SingleCellExperiment` must be correctly
named. The `cytomapper` package supports the visualization of up to 6 channels
and displays a combined image. However, in the case of displaying expression on
segmentation mask, the user should not display too many features.
See `?plotCells` for examples.

## Changing the assay slot

To visualize differently transformed counts, the `plotCells` function allows
setting the `exprs_values` parameter. In the toy dataset, the
`assay(pancreasSCE, "exprs")` slot contains the arcsinh-transformed raw
expression counts.

```{r exprs_values-cells}
plotCells(pancreasMasks, object = pancreasSCE, 
            img_id = "ImageNb", cell_id = "CellNb", 
            colour_by = c("SMA", "CD44"),
            exprs_values = "exprs")
```

## Outlining

The user can furthermore outline cells and specify the metadata entry to outline
cells by. See the previous [Outlining](#Outlining) section and `?plotCells` 
for examples.

## Subsetting

The user can further subset the images internally for visualization by setting
the `subset_images` parameter. Images can be subsetted by image ID, entry name
or simply by indexing. In the toy dataset, image IDs are numeric entries and
therefore the `subset_images` parameter behaves the same for image ID subsetting
and subsetting via integer indexing.
For examples, please see the previous [Subsetting](#subsetting) section.

## Adjusting the colour {#adjustingColour}

The user can also customize the colours for selected features and metadata. The
`colour` parameter takes a named `list` in which names correspond to the entries
to `colour_by` and/or `outline_by`. To colour continous features such as
expression or continous metadata entries (e.g. cell area), at least two colours
for interpolation need to be provided. These colours are passed to the
`colorRampPalette` function for interpolation. To colour discrete entries, one
colour per entry needs to be specified in form of a named vector.

```{r colour-cells}
plotCells(pancreasMasks, object = pancreasSCE, 
            img_id = "ImageNb", cell_id = "CellNb", 
            colour_by = c("SMA", "CD44"),
            outline_by = "CellType",
            colour = list(SMA = c("black", "red"),
                        CD44 = c("black", "white"),
                        CellType = c(celltype_A = "blue", 
                                    celltype_B = "green",
                                    celltype_C = "yellow")))
```

# Customisation

The next sections explain different ways to customise the visual output of the
`cytomapper` package. To find more details on additional parameters that can be
set to customise the display, refer to `?'plotting-param'`.

## Subsetting the SingleCellExperiment object

The `cytomapper` package matches cells contained in the `SingleCellExperiment`
to objects contained in the `CytoImageList` segmentation masks object via cell
identifiers. These are integer values, which are unique to each object per
image.

By matching these IDs, the user can subset the `SingleCellExperiment` object and
therefore only visualize the cells retained in the object:

```{r subsetting-SCE}
cur_sce <- pancreasSCE[,colData(pancreasSCE)$CellType == "celltype_A"]
plotCells(pancreasMasks, object = cur_sce, 
            img_id = "ImageNb", cell_id = "CellNb", 
            colour_by = "CellType",
            colour = list(CellType = c(celltype_A = "red")))
```

## Background and missing colour

The background of a segemntation mask is defined by the value `0`. To change the
background colour, the `background_colour` parameter can be set. Furthermore,
cells that are not contained in the `SingleCellExperiment` object can be
coloured by setting `missing_colour`. For an example, see figure
\@ref(fig:customization).

## Scale bar and image title

Depending on the cells' and background colour, the scale bar and image title are
not visible. To change the visual display of the scale bar, a named list can be
passed to the `scale_bar` parameter. The list should contain one or multiple of
the following entries: `length`, `label`, `cex`, `lwidth`, `colour`, `position`,
`margin`, `frame`. For a detailed explanation on the individual entries, please
refer to the `scale_bar` section in `?'plotting-param'`.

**Of note:** By default, the length of the scale bar is defined in number of
pixels. Therefore, the user needs to know the length (e.g. in $\mu{m}$) to label
the scale bar correctly.

The image titles can be set using the `image_title` parameter. Also here, the
user needs to provide a named list with one or multiple of follwing entries:
`text`, `position`, `colour`, `margin`, `font`, `cex`. The entry to `text` needs
to be a character vector of the same length as the `CytoImageList` object.

Plotting of the scale bar and image title can be suppressed by setting the
`scale_bar` and `image_title` parameters to `NULL`.

For an example, see figure \@ref(fig:customization).

## Legend

By default, the legend all all its contents are adjusted to the size of the
largest image in the `CytoImageList` object. However, legend features can be
altered by setting the `legend` parameter. It takes a named list containing one
or multiple of the follwoing entries: `colour_by.title.font`,
`colour_by.title.cex`, `colour_by.labels.cex`, `colour_by.legend.cex`,
`outline_by.title.font`, `outline_by.title.cex`, `outline_by.labels.cex`,
`outline_by.legend.cex`, `margin`. For detailed explanation on the individual
entries, please refer to the `legend` parameter in `?'plotting-param'`.

For an example, see figure \@ref(fig:customization).

## Setting the margin between images

To enhance the display of individual images, the `cytomapper` package provides
the `margin` parameter.

The `margin` parameter takes a single numeric indicating the gap (in pixels)
between individual images.

For an example, see figure \@ref(fig:customization).

## Scale the feature counts

By default, features are scaled to the minimum and maximum per channel. This
behaviour facilitates visualization but does not allow the user to visually
compare absolute expression counts across channels. The default behaviour can be
suppressed by setting `scale = FALSE`.

In this case, counts are linearly scaled to the minimum and maximum across all
channels and across all displayed images.

For an example, see figure \@ref(fig:customization).

## Image interpolation

By default, colours are interpolated between pixels (see `?rasterImage` for
details). To suppress this default behaviour, the user can set `interpolate =
FALSE`.

```{r customization, fig.cap = "Plot customization example."}
plotCells(pancreasMasks, object = pancreasSCE, 
            img_id = "ImageNb", cell_id = "CellNb", 
            colour_by = "SMA",
            outline_by = "CellType",
            background_colour = "white",
            missing_colour = "black",
            scale_bar = list(length = 30,
                            label = expression("30 " ~ mu * "m"),
                            cex = 2,
                            lwidth = 10,
                            colour = "cyan",
                            position = "bottomleft",
                            margin = c(5,5),
                            frame = 3),
            image_title = list(text = c("image_1", "image_2", "image_3"),
                            position = "topleft",
                            colour = "cyan",
                            margin = c(2,10),
                            font = 3,
                            cex = 2),
            legend = list(colour_by.title.font = 2,
                            colour_by.title.cex = 1.2,
                            colour_by.labels.cex = 0.7,
                            outline_by.legend.cex = 0.3,
                            margin = 10),
            margin = 2)
```

# Returning plots and images

The user has the option to save the generated plots (see next section) or to get
the plots and/or coloured images returned. If `return_plot` and/or
`return_images` is set to `TRUE`, `cytomapper` returns a list object with one or
two entries: `plot` and/or `images`.

The `display` parameter supports the entries `display = "all"` (default), which
displays images in a grid-like fashion and `display = "single"`, which display
images individually:

If the `return_plot` parameter is set to `TRUE`, `cytomapper` internally calls
the `recordPlot` function and returns a plot object. The user can additionally
set `display = "single"` to get a list of plots returned.

If the `return_images` parameter is set to `TRUE`, `cytomapper` returns a
`SimpleList` object containing three-colour (red, green. blue) `Image` objects.

```{r return_plot}
cur_out <- plotPixels(pancreasImages, colour_by = c("H3", "SMA", "CD44"), 
                return_plot = TRUE, return_images = TRUE, display = "single")
```

The returned plot objects now allows the plotting of individual images:

```{r return_plot-2}
cur_out$plot$A02_imc
```

Furthermore, the user can directly plot the coloured images from the returned
`SimpleList` object:

```{r return_images}
plot(cur_out$images$D01_imc)
```

However, when plotting solely the coloured images, the image title and scale bar
will be lost.

# Saving images

Finally, the user can save the plot by specifying `save_plot`. The `save_plot`
entry takes a list of two entries: `filename` and `scale`. The `filename` should
be a character representing a valid file name ending with `.png`, `.tiff` or
`.jpeg`. The `scale` entry controls the resolution of the image (see
`?"plotting-param"` for help). Increasing the scale parameter will increase the
resolution of the final image.

When setting `display = "single"`, the `cytomapper` package will save individual
images in individual files.

# Acknowledgements

We want to thank the [Bodenmiller laboratory](http://www.bodenmillerlab.org) for
feedback on the package and its functionality.

# Contributions

Nicolas created the first version of `cytomapper` (named `IMCMapper`). Nils and
Nicolas implemented and maintain the package.

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

# References

