---
title: "LoadPancreasData"
author: "Nicolas Damond"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = file.path("..", "extdata"))
```


# Script to download single-cell data from the pancreas IMC dataset and format the data as a SingleCellExperiment object
- Publication: [Damond et al. A Map of Human Type 1 Diabetes Progression by Imaging Mass Cytometry. Cell Metab. 2019 Mar 5;29(3):755-768](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6821395)
- Dataset: [accessible from Mendeley Data](http://dx.doi.org/10.17632/cydmwsfztj.2)

```{r load-packages, message=FALSE}
library(S4Vectors)
library(SingleCellExperiment)
library(cytomapper)
```

# Read-in single-cell data
Here, a subset of single-cell data, corresponding to 100 images from the full dataset is loaded

! The updated Mendeley dataset is not published yet, can be temporarily accessed from https://data.mendeley.com/datasets/cydmwsfztj/draft?a=e02ab17c-c014-4338-b379-101424e1c64d 
```{r download-cell-data}
# Download the zipped folder image and unzip it
# url.cells <- ("...")
# download.file(url.cells, destfile = "CellSubset.zip")
# unzip("CellSubset.zip")
# file.remove("CellSubset.zip")

# (Temporary code to unzip the data, will be removed when the Mendeley dataset is public)
fn.cells <- ("/home/nicolasd/Data/CM2017/mendeley/CellSubset.zip")
unzip(fn.cells)

# Read-in the data
cells <- read.csv("CellSubset.csv", stringsAsFactors = FALSE)

# Order the dataset by ImageNumber and ObjectNumber
cells <- cells[order(cells$ImageNumber, cells$ObjectNumber), ]
```

# Read-in Image metadata
```{r download-image-metadata}
# Download the zipped folder image and unzip it
# url.image <- ("...")
# download.file(url.image, destfile = "Image.zip")
# unzip("Image.zip")
# file.remove("Image.zip")

# (Temporary code to unzip the data, will be removed when the Mendeley dataset is public)
fn.images <- ("/home/nicolasd/Data/CM2017/mendeley/Image.zip")
unzip(fn.images)

# Read-in the data
image <- read.csv("All_Image.csv", stringsAsFactors = FALSE)
```

# Read-in cell type information
```{r download-celltypes}
# Download the zipped folder image and unzip it
# url.celltypes <- ("...")
# download.file(url.celltypes, destfile = "CellTypes.zip")
# unzip("CellTypes.zip")
# file.remove("CellTypes.zip")

# (Temporary code to unzip the data, will be removed when the Mendeley dataset is public)
fn.celltypes <- ("/home/nicolasd/Data/CM2017/mendeley/CellTypes.zip")
unzip(fn.celltypes)

# Read-in the data
celltypes <- read.csv("CellTypes.csv", stringsAsFactors = FALSE)
```

# Read-in donor information
```{r download-donor-info}
# Download the zipped folder image and unzip it
# url.donors <- ("...")
# download.file(url.donors, destfile = "Donors.zip")
# unzip("Donors.zip")
# file.remove("Donors.zip")

# (Temporary code to unzip the data, will be removed when the Mendeley dataset is public)
fn.donors <- ("/home/nicolasd/Data/CM2017/mendeley/Donors.zip")
unzip(fn.donors)

# Read-in the data
donors <- read.csv("Donors.csv", stringsAsFactors = FALSE)
```

# Load relevant cell-specific metadata
```{r cell-metadata}
cell.metadata <- DataFrame(ImageNumber = cells$ImageNumber,
                           CellNumber = cells$ObjectNumber,
                           Pos_X = cells$Location_Center_X,
                           Pos_Y = cells$Location_Center_Y,
                           ParentIslet = cells$Parent_Islets,
                           ClosestIslet = cells$Parent_ExpandedIslets,
                           Area = cells$AreaShape_Area,
                           NbNeighbours = cells$Neighbors_NumberOfNeighbors_3)
```

# Load relevant image-specific metadata
```{r image-metadata}
image.metadata <- DataFrame(ImageNumber = image$ImageNumber,
                            ImageFullName = image$FileName_CleanStack,
                            slide = image$Metadata_Slide,
                            width = image$Width_CleanStack,
                            height = image$Height_CleanStack)
```

# Merge cell and image metadata
```{r merge-metadata}
cell.metadata <- merge(cell.metadata, image.metadata, by="ImageNumber")
```

# Add image names
This information is used by SingleCellMapper to match single-cell data with images and masks
```{r add-image-names}
cell.metadata$ImageName <- sub("_a0_full_clean.tiff", "", cell.metadata$ImageFullName)
```

# Import cell types
```{r import-cell-types}
# Add cell ids to cell metadata (format: "ImageName_CellNumber")
cell.metadata$id <- paste(cell.metadata$ImageName, cell.metadata$CellNumber, sep="_")

# Merge cell metadata and cell type information
cell.metadata <- merge(cell.metadata,
                       celltypes[, c("id", "CellCat", "CellType")],
                       by="id")
```

# Import donor metadata
```{r import-donors-info}
cell.metadata <- merge(cell.metadata, donors, by="slide")
```

# Order the cell metadata dataset and add rownames
```{r cell-rownames}
# Rows are ordered by ImageNumber and CellNumber
cell.metadata <- cell.metadata[order(cell.metadata$ImageNumber, cell.metadata$CellNumber), ]

# Cell ids are used as row names
rownames(cell.metadata) <- cell.metadata$id
```

# Load panel data
The panel contains antibody-related metadata.
The channel-mass file is used to match panel information and image stack slices.
```{r load-panel}
# Import panel
# url.panel <- ("...")
# download.file(url.panel, destfile = "panel.csv")
# panel <- read.csv(panel.csv)

# Import channel-mass file
# url.channelmass <- ("...")
# download.file(url.channelmass, destfile = "ChannelMass.csv")
# panel <- read.csv(ChannelMass.csv)

# Temporary code
panel <- read.csv("/home/nicolasd/Data/CM2017/mendeley/Panel.csv")
channel.mass <- read.csv("/home/nicolasd/Data/CM2017/mendeley/ChannelMass.csv", header=F)
```

# Select relevant channels and match them with image stack slices
```{r arrange-panel}
# Match panel and stack slice information
panel <- panel[panel$full == 1,]
panel <- panel[match(channel.mass[,1], panel$MetalTag),]

# Add short protein names as panel rownames
rownames(panel) <- panel$shortname
```

# Load single cell measurements
Here, we import the mean intensity per cell
```{r select-counts}
cur_counts <- cells[, grepl("Intensity_MeanIntensity_CleanStack", colnames(cells))]
```

# Reorder the counts channels (based on channel number)
```{r reorder-counts}
channelNumber <- as.numeric(sub("^.*_c", "", colnames(cur_counts)))
cur_counts <- cur_counts[, order(channelNumber, decreasing = FALSE)]
```

# Create the SingleCellExperiment (SCE) object
```{r create-SCE}
sce <- SingleCellExperiment(assays = list(counts = t(as.matrix(cur_counts))))
```

# Add transformed counts as a new assay
exprs = asinh-transformed counts
```{r transform-counts}
assay(sce, "exprs") <- asinh(counts(sce)/1)
```

# Set rownames
```{r rownames-SCE}
rownames(sce) <- rownames(panel)
colnames(sce) <- rownames(cell.metadata)
```

# Store metadata in the SCE object
```{r store-metadata}
colData(sce) <- cell.metadata
rowData(sce) <- panel
sce
```

# Save SCE
```{r save-SCE}
saveRDS(sce, "pancreas_sce.rds")
```

# Delete unneeded CSV files from the extdata directory
```{r clean-up, message=F}
file.remove("CellSubset.csv", "All_Image.csv", "CellTypes.csv", "Donors.csv", "Panel.csv", "ChannelMass.csv")
```

